---
  title: "Final Project"
author: "Team 6"
output:
  pdf_document: default
html_document:
  df_print: paged
---

*Libraries*

```{r}
# Load necessary libraries
library(readxl)
library(writexl)
library(corrplot)
library(dplyr)
library(caret)
library(glmnet)
library(knitr)
library(boot)
library(smotefamily)
library(RANN)
library(ggplot2)
library(tidyr)
library(forcats)
```

*Data cleaning*

```{r include_data_readme, comment=''}
# Load the data
data <- read_excel("data/data_toshare.xlsx")


# Display the first few rows to understand the data structure
cat("First few rows of the original data:\n")
print(head(data))

# Calculate the number of rows and columns before cleaning
initial_rows <- nrow(data)
initial_cols <- ncol(data)

# Remove specified columns
columns_to_remove <- c("ResponseId", "add", "lang", "sday", "smonth", "syear","Progress" )
data <- data[, !(names(data) %in% columns_to_remove)]

# Remove columns with more than 50% null values
threshold_col <- 0.5 * nrow(data)
cleaned_data <- data[, colSums(is.na(data)) <= threshold_col]

# Replace nulls in numeric columns with the median value
numeric_columns <- sapply(cleaned_data, is.numeric)
cleaned_data[, numeric_columns] <- lapply(cleaned_data[, numeric_columns], function(col) {
  col[is.na(col)] <- median(col, na.rm = TRUE)
  return(col)
})

# Remove rows with more than 50% null values
threshold_row <- 0.5 * ncol(cleaned_data)
cleaned_data <- cleaned_data[rowSums(is.na(cleaned_data)) <= threshold_row, ]

# Calculate the number of rows and columns after cleaning
final_rows <- nrow(cleaned_data)
final_cols <- ncol(cleaned_data)

# Display the number of rows and columns before and after cleaning
cat("Number of rows before cleaning:", initial_rows, "\n")
cat("Number of rows after cleaning:", final_rows, "\n")
cat("Number of columns before cleaning:", initial_cols, "\n")
cat("Number of columns after cleaning:", final_cols, "\n")

# Define the statistic function for bootstrapping
mean_lifesat <- function(data, indices) {
  d <- data[indices, ]  # allows boot to select sample
  return(mean(d$lifesat))
}
print(head(cleaned_data))

data <- cleaned_data
```


*Correlation matrix*

```{r}

# Identify numerical and categorical columns
num_cols <- sapply(data, is.numeric)
cat_cols <- sapply(data, is.character)

# Handle missing values
# For numerical columns, impute with mean
data[, num_cols] <- lapply(data[, num_cols], function(x) ifelse(is.na(x), mean(x, na.rm = TRUE), x))

# For categorical columns, impute with mode
data[, cat_cols] <- lapply(data[, cat_cols], function(x) {
  mode_value <- names(sort(table(x), decreasing = TRUE))[1]
  ifelse(is.na(x), mode_value, x)
})

# Convert categorical variables to factors
data[, cat_cols] <- lapply(data[, cat_cols], as.factor)

# Remove columns with zero variance
non_zero_var_cols <- sapply(data, function(x) var(x, na.rm = TRUE) != 0)
data <- data[, non_zero_var_cols]

# Calculate the correlation matrix
cor_matrix <- cor(data[, sapply(data, is.numeric)], use = "complete.obs")

# Plot the correlation matrix
corrplot(cor_matrix, method = "color", tl.cex = 0.7, tl.col = "black")

# Select the features that have a high correlation with the target variable 'lifesat'
target_cor <- cor_matrix[,"lifesat"]
high_cor_features <- names(target_cor[abs(target_cor) > 0.3])  # Adjust the threshold as needed
relevant_features <- target_cor[high_cor_features]

# Convert the relevant features to a data frame
relevant_features_df <- as.data.frame(relevant_features)
kable(relevant_features_df, format = "markdown", align = "c")

# From the relevant features the correlation matrix returns we choose the most relevent for our model

most_relevant_features <- c("spa", "sph", "partner", "qol", "cesd", "gai", "lon", "int_homophobia", "fdistress")

```

*Creating model*

```{r}

# For categorical columns, impute with mode
cat_cols <- sapply(data, is.character)
data[cat_cols] <- lapply(data[cat_cols], function(x) {
  mode_value <- names(sort(table(x), decreasing = TRUE))[1]
  ifelse(is.na(x), mode_value, x)
})

# Convert categorical variables to factors
data[cat_cols] <- lapply(data[cat_cols], as.factor)


# Select the features that exist in the data
X <- data %>% select(all_of(most_relevant_features))
y <- data$lifesat

# Combine X and y for SMOTE
data_combined <- cbind(X, lifesat = y)


# Split the data into training and testing sets
set.seed(42)
trainIndex <- createDataPartition(y, p = .8, list = FALSE, times = 1)
X_train <- X[trainIndex, ]
X_test <- X[-trainIndex, ]
y_train <- y[trainIndex]
y_test <- y[-trainIndex]

# Standardize the numerical features
preProcValues <- preProcess(X_train, method = c("center", "scale"))
X_train <- predict(preProcValues, X_train)
X_test <- predict(preProcValues, X_test)

# Ridge regression
ridge_model <- cv.glmnet(as.matrix(X_train), y_train, alpha = 0)
ridge_best_lambda <- ridge_model$lambda.min
ridge_predictions <- predict(ridge_model, s = ridge_best_lambda, newx = as.matrix(X_test))
ridge_rmse <- sqrt(mean((ridge_predictions - y_test)^2))

# Lasso regression
lasso_model <- cv.glmnet(as.matrix(X_train), y_train, alpha = 1)
lasso_best_lambda <- lasso_model$lambda.min
lasso_predictions <- predict(lasso_model, s = lasso_best_lambda, newx = as.matrix(X_test))
lasso_rmse <- sqrt(mean((lasso_predictions - y_test)^2))

# Print the results
print(paste("Ridge RMSE:", ridge_rmse))
print(paste("Lasso RMSE:", lasso_rmse))
print(paste("Best lambda for Ridge:", ridge_best_lambda))
print(paste("Best lambda for Lasso:", lasso_best_lambda))

# The results shows that the lasso model is slightly better so we choose it
```
*Display of Lasso model and error evaluation*

```{r}

# Print Lasso model details
print("Coefficients for Lasso model:")
print(coef(lasso_model, s = lasso_best_lambda))

# Print Lasso model details using knitr
lasso_coefficients <- as.matrix(coef(lasso_model, s = lasso_best_lambda))
lasso_coefficients_df <- data.frame(Feature = rownames(lasso_coefficients), Coefficient = lasso_coefficients[, 1])
kable(lasso_coefficients_df, format = "markdown", align = "c")

# Lasso regression
lasso_model <- cv.glmnet(as.matrix(X_train), y_train, alpha = 1)
lasso_best_lambda <- lasso_model$lambda.min
lasso_predictions <- predict(lasso_model, s = lasso_best_lambda, newx = as.matrix(X_test))
lasso_rmse <- sqrt(mean((lasso_predictions - y_test)^2))
lasso_mse <- mean((lasso_predictions - y_test)^2)
lasso_mae <- mean(abs(lasso_predictions - y_test))
lasso_r2 <- 1 - sum((lasso_predictions - y_test)^2) / sum((y_test - mean(y_test))^2)

# Print the results
print(paste("RMSE:", lasso_rmse))
print(paste("MSE:", lasso_mse))
print(paste("MAE:", lasso_mae))
print(paste("R²:", lasso_r2))

print(paste("Best lambda for Lasso:", lasso_best_lambda))

```

*Visualization*

*Correlation between Homophobia, Mental Health, and Environmental Support*
```{r}
# Assuming 'selected_data' is your cleaned and preprocessed DataFrame
# Select relevant columns

# Create a combined score for mental health (negative direction: higher score = worse mental health)
selected_data <- data %>%
  mutate(mental_health = cesd + gai + lon,
         environment_support = bio_stable + sp_stable + choice_stable)

# Discretize the combined environmental support into bins for faceting
selected_data <- selected_data %>%
  mutate(environment_support_bins = cut(environment_support, 
                                        breaks = 4, 
                                        labels = c("Low", "Medium-Low", "Medium-High", "High")))

# Define LGBTQ+ color palette
lgbtq_colors <- c("red", "orange", "yellow", "green", "blue", "purple")

# Plot the relationships with facets based on combined environmental support
ggplot(selected_data, aes(x = int_homophobia, y = mental_health)) +
  geom_point(aes(color = environment_support)) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  facet_wrap(~ environment_support_bins) +
  scale_color_gradientn(colors = lgbtq_colors) +
  labs(title = "Correlation between Homophobia, Mental Health, and Environmental Support",
       x = "Internalized Homophobia",
       y = "Mental Health issues",
       color = "Environmental Support") +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    strip.text = element_text(size = 14),
    axis.text.x = element_text(size = 10),
   
    panel.background = element_rect(fill = "#F6F5F2"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "white"),
    panel.grid.minor = element_line(color = "white")
  )

```

*Impact of Family Closeness on Quality of Life (not in the report)*
```{r}

Closeness_data <- data %>%
  mutate(bio_close = case_when(
    bio_close == 1 ~ "Strongly Disagree",   
    bio_close == 2 ~ "Disagree", 
    bio_close == 3 ~ "Neutral",
    bio_close == 4 ~ "Agree",
    bio_close == 5 ~ "Strongly Agree"
  )) %>%
  mutate(bio_close = factor(bio_close, levels = c("Strongly Disagree", "Disagree", "Neutral", "Agree", "Strongly Agree")))

# Plotting
ggplot(data, aes(x = bio_close, y = qol)) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Impact of Family Closeness on Quality of Life",
       x = "Family Closeness",
       y = "Quality of Life") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())


```

*Impact of Closet Status on Life Satisfaction*
```{r}
# Define pastel LGBTQ+ pride colors
pastel_lgbtq_colors <- c("#ffc09f", "#ffee93", "#adf7b6", "#B3D9FF")

# Ensure the closet column is a factor and update the levels
closet_data <- data %>%
  mutate(closet = factor(closet, levels = 1:4, labels = c(
    "Mostly Out",
    "Partially Out",
    "Barely Out",
    "Not Out"
  )))

# Summarize the data to get mean Life Satisfaction for each level of closet
closet_summary <- closet_data %>%
  group_by(closet) %>%
  summarize(mean_lifesat = mean(lifesat, na.rm = TRUE))

# Bar plot with pastel LGBTQ+ pride colors
ggplot(closet_summary, aes(x = closet, y = mean_lifesat, fill = closet)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  scale_fill_manual(values = pastel_lgbtq_colors) +
  labs(title = "Impact of Closet Status on Life Satisfaction",
       x = "Closet Status",
       y = "Mean Life Satisfaction") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())


```
*relationship between family Support and depression/anxiety*
```{r}

rainbow_palette <- c("Depression" = "#eac4d5", "Anxiety" = "#95b8d1")

# Social Support and Depression/Anxiety
social_support_plot <- data %>%
  select(bio_suprec, bio_supgv, cesd, gai) %>%
  gather(key = "Measure", value = "Score", cesd, gai) %>%
  mutate(Measure = recode(Measure, cesd = "Depression", gai = "Anxiety")) %>%
  ggplot(aes(x = bio_suprec, y = Score, color = Measure)) +
  geom_smooth(method = "lm", se = FALSE, size = 1.5) +
  scale_color_manual(values = rainbow_palette) +
  labs(title = "Family Support and Depression/Anxiety",
       x = "Receiving Help from Family",
       y = "Depression/Anxiety Scores",
       color = "Measure") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

# Display the plot
print(social_support_plot)


```
